---
title: "Reproducible Research Assignment 1"
author: "Rajeev Kumar"
date: "November 14, 2015"
output: html_document
---

##  Loading and preprocessing the data

```{r echo=TRUE, cache=TRUE}
    suppressMessages(library(dplyr))
    url<-"./repdata-data-activity/activity.csv"
    actDF<-read.csv(url,stringsAsFactors = FALSE)
    actDF<-tbl_df(actDF)
    actDF

```
##  What is mean total number of steps taken per day ?  the missing values can be ignored

### 1.  Reporting the total number of steps taken per day

```{r echo=TRUE}
    suppressMessages(library(dplyr))
    by_date <-group_by(actDF,date)
    by_date<-arrange(by_date,date,steps)
    summary_by_day<-summarize(by_date,number_Of_Steps=sum(steps,na.rm=TRUE),
                    mean_of_steps=mean(as.numeric(steps),na.rm=TRUE),
                    median_of_steps=median(as.numeric(steps),na.rm=TRUE))
    summary_by_day[,c(1,2)]
    
``` 

### 2.  Histogram of total steps taken per day

```{r echo=TRUE}
    
    hist(as.numeric(summary_by_day$number_Of_Steps),main="Total steps per day ",xlab="Steps taken per day" ,col="red")

```    

### 3.  Reporting the mean and median of the total number of steps taken per day

```{r echo=TRUE}
    suppressMessages(library(dplyr))
    summary_by_day<-summarize(by_date,number_Of_Steps=sum(steps,na.rm=TRUE),
                    mean_of_steps=mean(as.numeric(steps),na.rm=TRUE),
                    median_of_steps=median(as.numeric(steps),na.rm=TRUE))
    
    summary_by_day[,c(1,3,4)]
```

##  What is mean total number of steps taken per day?
### 1.  Time series plot

```{r echo=TRUE, cache=TRUE}
    suppressMessages(library(dplyr))
    group_by_interval<-group_by(actDF,interval)
    group_by_interval<-arrange(group_by_interval,interval)
    summary_by_interval<-summarize(group_by_interval,mean_of_steps=mean(as.numeric(steps),na.rm=TRUE))
    plot(summary_by_interval$interval,as.numeric(summary_by_interval$mean_of_steps),xlab="",ylab="", type="l")
    title(xlab="Interval",ylab="Mean Of Steps",line=3)
    
```  

### 2.  Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r echo=TRUE}
    suppressMessages(library(dplyr))
    max_steps_by_interval<-arrange(summary_by_interval,desc(mean_of_steps))
    interval_with_max_step_val<- max_steps_by_interval[1,"interval"][[1]]
    interval_with_max_step_val
```    

##  Missing Values

### 1.  Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r echo=TRUE}
    steps<-actDF[,"steps"]
    totalMissingValues<-length(steps[is.na(steps)])
    totalMissingValues
```
### 2.  Filling missing values based upon mean of that day and creating the new dataset
```{r echo=TRUE}
    vector_of_days_mean<-summary_by_day[,"mean_of_steps"][[1]]
    totalDays<-61
    totReadingPerDay<-288
    updDF<-actDF
    num<-0
    
    for(i in 1:totalDays){
        for(j in 1:totReadingPerDay){
            num<-num+1
            #if(is.na(updDF[num,"steps"])[[1]]){
            if(is.na(updDF[num,"steps"])){
                if(is.na(vector_of_days_mean[i])){
                    updDF[num,"steps"]<-0
                }else{
                    ##Assigning mean of the day to missing values of that day
                    updDF[num,"steps"]<-vector_of_days_mean[i]
                    
                }
            }
            
        }
        
    }
    
    updDF
```    



### 3.  
#### A. Make a histogram of the total number of steps taken each day based on new dataset.
```{r echo=TRUE,cache=TRUE}
    by_date1 <-group_by(updDF,date)
    by_date1<-arrange(by_date1,date,steps)
    summary_by_day1<-summarize(by_date1,number_Of_Steps=sum(steps,na.rm=TRUE),
                    mean_of_steps=mean(as.numeric(steps),na.rm=TRUE),
                    median_of_steps=median(as.numeric(steps),na.rm=TRUE))
    
    hist(as.numeric(summary_by_day$number_Of_Steps),main="Total steps per day ",xlab="Steps taken per day" ,col="blue")

```

#### B. Report the mean and median total number of steps taken per day. 
```{r echo=TRUE}
    summary_by_day1
    
```
#### C. Do these values differ from the estimates from the first part of the assignment? 
##### a.    A comparision of mean steps per day is as shown below 
```{r echo=TRUE}
    mean_with_missing_vals<-summary_by_day[,c(1,3)]
    mean_without_missing_vals<-summary_by_day1[,c(3)]
    mean_DF<-data.frame(mean_with_missing_vals,mean_without_missing_vals)
    mean_DF
```
##### b.    A comparision of median steps per day is as shown below 
```{r echo=TRUE}
    median_with_missing_vals<-summary_by_day[,c(1,2,4)]
    median_without_missing_vals<-summary_by_day1[,c(4)]
    median_DF<-data.frame(median_with_missing_vals,median_without_missing_vals)
    median_DF
```

#### D. What is the impact of inputing missing data on the estimates of the total daily number of steps?
```{r echo=TRUE}
    comp_DF<-data.frame(summary_by_day[,c(1,2)],summary_by_day[,c(2)])
    comp_DF
```

##  Are there differences in activity patterns between weekdays and weekends?

### 1.  Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a             given date is a weekday or weekend day.
```{r echo=TRUE, cache=TRUE}
    updDF<-mutate(updDF,day_of_week=weekdays(as.Date(date)))
   
    for(i in 1:17568){
       
            if(updDF[i,"day_of_week"] %in% c("Monday","Tuesday","Wednesday","Thursday")){
                updDF[i,"day_of_week"]<-"Weekday"
                
            }else{
                
                updDF[i,"day_of_week"]<-"Weekend"
            }
        
    }
    
    #setting factor variable with 2 levels weekday, weekend
    updDF$day_of_week<-as.factor(updDF$day_of_week)
    str(updDF)
   
```
### 2. Plot
```{r echo=TRUE}
     group_by_interval<-group_by(updDF,interval)
     group_by_interval<-arrange(group_by_interval,interval)
     summary_by_interval<-summarize(group_by_interval,mean_of_steps=mean(as.numeric(steps),na.rm=TRUE))
   
     library(lattice)
    xyplot(summary_by_interval$mean_of_steps ~summary_by_interval$interval|updDF$day_of_week,layout=c(1,2),type="l")
    
```    